Structuring Theories with Implicit Morphisms

### 2018-10-13 mmt.tex
Bands example

\textbf{Flat theories} are lists of declarations $c[:E][=e]$ where $E$ and $e$ are optional expressions.
We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$.
Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=P$, which establish theorem $F$ via proof $P$.
Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

@@

\paragraph{Well-Formedness}
The inference system to define well-formed modules, and expressions is not a primary interest of this paper.
Therefore, we only mention those aspects that are relevant later on.

However, one detail will be critical later on:
the inference rules for the well-formedness of expressions include the following base case for constants:
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c:E} \tb \rul{c[:E]=e \minn \flt{T}}{\vdash_T c=e}\]
Correspondingly, the definition of the homomorphic extension of a morphism includes the following base case for constants:
\[M(c) = e \mwhere c:= e\in\flt{M}\]

These base cases introduce a mutual recursion between well-formedness and flattening:
the well-formedness of a declaration in a theory depends on the flattening of all preceding declarations;
and the well-formedness of an assignment in a morphism depends on the homomorphic extension of the morphism obtained by flattening of all preceding assignments.
Vice versa, well-formedness is a precondition for defining the flattening.
It is desirable to define well-formedness independently of flattening.
But the mutual recursion makes sense from an implementation perspective: typical tools for formal systems will first parse, check, and flatten the first declaration, then continue with the next declaration.
In fact, often the checking of well-formedness is as expensive as flattening anyway.

Therefore, we will make flattening a partial function, i.e., $\flt{X}$ is defined iff the module $X$ is well-formed.

@@

@@

\begin{definition}[Theory]\label{def:theory}
A \textbf{theory} $T$ is of the form
\begin{grammar}
TDec     & t=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
Dec      & n\opt{:o}\opt{=o}     & \text{constant declaration}\\
c        & t?n                   & \text{qualified constant identifiers} \\
o        & c \alt \ldots         & \text{expressions built from constants} \\
T        & t                     & \text{theory expressions}
\end{grammar}

@@

@@

@@

\begin{example}[Includes]\label{syn:incl}
We extend the grammar with
\begin{grammar}
  Dec & \icl{S} & \text{include a theory into a theory} \\
  Ass & \icl{M} & \text{include a morphism into a morphism}
\end{grammar}
\end{example}

@@

At this point, we only have one case for declarations $D$, namely constant declarations.
Their flattening is trivial:
 \[\fltd{\Sigma}{n[:E][=e]}=\{t?n[:E][=e]\}\]
where $t$ is the name of the containing theory.

Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
with the trivial base case
 \[\fltd{\sigma}{c:=e}=\{c:=o\}\]

@@

Finally, we define the semantics of the category structure:
\begin{example}[Category of Theories (continued from Ex.~\ref{syn:cat})]\label{sem:cat}
For $T\in\Thy$, we put $\id{T}\in\Mo{T}{T}$ and define
\[\flt{\id{T}}=\{c:=c \;|\; c\in\dom{\flt{T}}\}\]

For $R,S,T\in\Thy$ and $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we put $M;N\in\Mo{R}{T}$ and define
\[\flt{M;N}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c\in\dom{\flt{R}}\}\]
\end{example}

>>>>

\textbf{Flat theories} are lists of declarations $c:E[=e]$ where $E$ and $e$ are expression, and the latter is optional.
We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$ (see below for the definition of well-formed expressions).
Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=P$, which establish theorem $F$ via proof $P$.
Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

@@

\paragraph{Well-Formed Expressions}
The inference system to define well-formed expressions is not a primary interest of this paper.
Intuitively, the set $\Exp{T}$ of expressions over $T$ includes all syntax trees that can be formed using the constants from $\flt{T}$.
Moreover, well-formed modules may only use expressions that satisfy certain typing and equality judgments.
We refer to \cite{rabe:howto:14} for details.

The only aspect of the inference system for these judgments that is relevant to our purposes here is the rules for constants:
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
Here, in order to avoid case distinctions for the case where a definition is present or not, we use a combined typing+equality judgment: $e_1=e_2:E$ represents the conjunction of $e_1:E$, $e_2:E$, and $e_1=e_2$.
Correspondingly, the definition of the homomorphic extension of a morphism with domain $S$ includes the following case for constants:
\[M(c) = \cas{e\mifc (c:E) \in \flt{S}, (c:=e)\in\flt{M} \\ M(e) \mifc (c:E=e)\in \flt{S}}\]
Here if $c$ has a definiens in $S$, we expand it before applying $M$.%
\footnote{The \mmt tool also $c:e\in \flt{M}$ even if $(c:E=e')\in\flt{S}$ has a definition. In that case, \mmt checks $\vdash_T M(e')=e$ and puts $M(c)=e$. This is important for efficiency but not essential for our purposes here.}

Note that these base cases introduce a mutual recursion between well-formedness and flattening:
the well-formedness of a declaration in a theory depends on the flattening of all preceding declarations;
correspondingly, the well-formedness of an assignment in a morphism depends on the homomorphic extension of the morphism obtained by flattening of all preceding assignments.
Vice versa, well-formedness is a precondition for defining the flattening --- the definition of flattening may become nonsensical if applied to ill-formed modules.

It is desirable to define well-formedness independently of flattening.
But the mutual recursion makes sense from an implementation perspective: typical tools for formal systems first parse, check, and flatten a declaration entirely before moving on to the next declaration.
Moreover, often the checking of well-formedness is as difficult or expensive as flattening anyway.
In fact, inspecting practical systems with modular features (which often do not have a full rigorous formal specification) shows that our definition elegantly captures the essential commonalities between them.

Therefore, we will make flattening a partial function, i.e., $\flt{X}$ is undefined if the module $X$ is not well-formed.

@@

\paragraph{Logics in \mmt}
\ednote{explain that we use a fixed instance of MMT which adds LF and FOL; only relevant for examples}

@@

\begin{definition}[Theory]\label{def:theory}
A \textbf{theory} $T$ is of the form
\begin{grammar}
TDec     & t=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
Dec      & n:o\opt{=o}           & \text{constant declaration}\\
c        & t?n                   & \text{qualified constant identifiers} \\
o        & c \alt \ldots         & \text{expressions built from constants} \\
T        & t                     & \text{theory expressions}
\end{grammar}

@@

\begin{example}\label{ex:thgroup}
The (flat) theories $\cn{Group}$ and $\cn{DivGroup}$ from the introduction are:
\begin{mmtcode}
Group =
  universe	: type	 # U
  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
  unit		: U
  inverse : U ⟶ U   # 1 ⁻¹

  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z 
  axiom_right_unit	: ⊦ ∀[x] x ∘ unit ≐ x 
  axiom_left_unit	: ⊦ ∀[x] unit ∘ x ≐ x 
  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ unit
\end{mmtcode}
\begin{mmtcode}
DivisionGroup =
  universe : type  # U prec 1 
  div : U ⟶ U ⟶ U  # 1 / 2 
  unit : U  # e prec 1 
  // axioms ... 
\end{mmtcode}
Here we use \lstinline|# <not>| to indicate notations; these are part of the language supported by \mmt but omitted from the formal grammar above because we only need them here to present legible examples.

Moreover, we use a few constants from a type systems\ednote{finish}$\forall,\doteq$ etc.
\end{example}

@@

\begin{example}\label{ex:dg2g}
	We can now specify the morphisms between our two definitions of groups. The theory of groups as monoids with inverses is given in \autoref{syn:incl}, the theory $\cn{DivGroup}$ and the morphism $\cn{DG2G}$ are as follows:
\begin{mmtcode}
DG2G : DivisionGroup -> Group =
  universe = universe 
  div = [a,b] a ∘ (b⁻¹) 
  unit = unit
  // axioms ...
\end{mmtcode}
The morphism $\cn{DG2G}$ hence maps the universe of a division group to the universe of a group, division to the expression $\lambda a,b.\; a\circ b^{-1}$ and the witness $\cn{nonEmpty}$ of a division group to the unit $e$ of a group. Additionally, the view has to map the axioms of a division group to their \emph{proofs} in $\cn{Group}$ of their translated statements.

The reverse morphism $\cn{G2DG}$ is completely analogous, mapping $\cn{op}$, $\cn{unit}$ and $\cn{inverse}$ to their corresponding definitions in terms of the division operation $\cn{div}$.
\end{example}

@@

\begin{example}[Includes]\label{syn:incl}
We extend the grammar with
\begin{grammar}
  Dec & \icl{S} & \text{include a theory into a theory} \\
  Ass & \icl{M} & \text{include a morphism into a morphism}
\end{grammar}
This allows writing the theory $\cn{Group}$ as in \autoref{ex:thgroup} by extending a theory of monoids:
\begin{mmtcode}
Monoid =
  universe	: type		 # U 
  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
  unit		: U
  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z 
  axiom_right_unit	: ⊦ ∀[x] x ∘ unit ≐ x 
  axiom_left_unit	: ⊦ ∀[x] unit ∘ x ≐ x

Group =
  include Monoid
  inverse : U ⟶ U   # 1 ⁻¹ 
  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ unit
\end{mmtcode}
\end{example}

@@

At this point, we only have one case for declarations $D$, namely constant declarations.
Their flattening is trivial:
 \[\fltd{\Sigma}{n:E[=e]}=\{t?n:E[=e]\}\]
where $t$ is the name of the containing theory.

Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
with the trivial base case
 \[\fltd{\sigma}{c:=e}=\{c:=e\}\]

@@

Finally, we define the semantics of the category structure:
\begin{example}[Category of Theories (continued from Ex.~\ref{syn:cat})]\label{sem:cat}
For $T\in\Thy$, we put $\id{T}\in\Mo{T}{T}$ and define
\[\flt{\id{T}}=\{c:=c \;|\; c:E\in\dom{\flt{T}}\}\]

For $R,S,T\in\Thy$ and $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we put $M;N\in\Mo{R}{T}$ and define
\[\flt{M;N}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c:E\in\dom{\flt{R}}\}\]

Note that in both cases, we only have to consider constants $c:E$ without definiens.
\end{example}

### 2018-10-13 paper.tex
rewritten Sect. 1+2
update
Bands example

\usepackage{sembracket}
\usepackage{mytikz}
\usepackage{basics}
%\usepackage[llncs]{theorems} %only for now so that the environments are defined; replace with nice package for this later
\usepackage{local}

@@

@@

\input{wadtabstract}
\bibliographystyle{alpha}
\bibliography{../../../Program_Data/Latex/bib/rabe,../../../Program_Data/Latex/bib/systems,../../../Program_Data/Latex/bib/institutions,../../../Program_Data/Latex/bib/pub_rabe,../../../Program_Data/Latex/bib/historical}
\end{document}

@@

\section{Theories and Theory Morphism}
\input{mmt}

\section{Inclusion Morphisms}\label{sec:include}
\input{include}

@@

\section{Complex Theories and Morphisms}\label{sec:complex}
\input{complex}

@@

\bibliographystyle{alpha}
\bibliography{../../../Program_Data/Latex/bib/rabe,../../../Program_Data/Latex/bib/systems,../../../Program_Data/Latex/bib/institutions,../../../Program_Data/Latex/bib/pub_rabe,../../../Program_Data/Latex/bib/historical}

>>>>

\usepackage{macros/sembracket}
\usepackage{macros/mytikz}
\usepackage{macros/basics}
%\usepackage[llncs]{theorems} %only for now so that the environments are defined; replace with nice package for this later
\usepackage{local}

\input{macros/listings}

@@

\pagestyle{plain}

@@

@@

\section{Theories and Theory Morphism}\label{sec:mmt}
\input{mmt}

%\section{Inclusion Morphisms}\label{sec:include}
%\input{include}

@@

\section{Applications}
\input{appl}

\section{Extended Example}
\input{example}

%\section{Complex Theories and Morphisms}\label{sec:complex}
%\input{complex}

@@

\bibliographystyle{alpha}
\bibliography{../../../../frabe/Program_Data/Latex/bib/rabe,../../../../frabe/Program_Data/Latex/bib/systems,../../../../frabe/Program_Data/Latex/bib/institutions,../../../../frabe/Program_Data/Latex/bib/pub_rabe,../../../../frabe/Program_Data/Latex/bib/historical}
